Powering Hidden Markov Model by Generative Models

### 2019-08-10 manuscript/main.tex
adjust sections

\documentclass{article}

\usepackage{hyperref} %BUG if put after
\usepackage{tikz}
\usetikzlibrary{calc,shapes,positioning}
\usetikzlibrary{arrows}
\newcommand{\midarrow}{\tikz \draw[-triangle 90] (0,0) -- +(.1,0);}
% Be sure to use PDxF Latex
\pdfoutput=1

@@

\usepackage{url}
\usepackage{fullpage}
\usepackage{cite}
\usepackage{caption}
\usepackage{bm}
\newcommand{\ubar}[1]{\mkern2mu\underline{\mkern-2mu #1\mkern-2mu}\mkern2mu}
% \allowdisplaybreaks
\usepackage{mystyle}
\newcommand{\ubm}[1]{\ubar{\bm{#1}}}
\newcommand{\ubmr}[2]{\ubar{\bm{#1}}^{(#2)}}

@@

\hypersetup{  
  bookmarks=true,
  backref=true,
  pagebackref=false,
  colorlinks=true,
  linkcolor=blue,
  citecolor=red,
  urlcolor=blue,
  pdftitle={Generative Model HMM},
  pdfauthor={Dong Liu},
  pdfsubject={}
}

\title{Powering Hidden Markov Model by Generative Models}
\author{
  Firsthand Scientists
}

@@

\maketitle
\section{Notation}
Time is indexed by subscript and sequence is denoted by underline. $\bm{x}_t$ is signal at time $t$. The sequential time is denoted by $\ubar{\bm{x}} = \left[ \bm{x}_1,
  \cdots, \bm{x}_T\right]^{\intercal}$, where $[\cdot]^{\intercal}$ means transpose and $T$ is the length of the sequence.  Sequential signal or clip uses underline notation and is indexed by superscript, for instance $\ubar{\bm{x}}^{(r)}$ means the $r$-th sequential signal, where $r = 1, 2, \cdots, R.$, and $\ubmr{x}{r} = \left[ \bmtr{x}{1}{r}, \bmtr{x}{2}{r}, \cdots, \bmtr{x}{T^{(r)}}{r} \right]$ with length ${T^{(r)}}$. Note different sequential signal $\ubmr{x}{r}$ could have different lengths.

The hypothesis of Hidden Markov Model (HMM): $\Hh := \{\bm{H} | \{\Ss, \bm{q}, A, p(\bm{x}|{s}; \bm{\Phi})\}$,
\begin{itemize}
\item $\Ss$ is the set of states of HMM $\bm{H}$;
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ initial distribution of HMM $\bm{H}$ with $|\Ss|$ is cardinality of $\Ss$, $q_k = p(s=k)$ for random state variable $s$.
\item $A$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$.
\item Observable signal density $p({\bm{x}}|{s};\bm{\Phi})$ given hidden state sequence, where $\bm{\Phi}$ is the parameter set that defines this conditional probabilistic model.
\end{itemize}

\section{Problem Statement}
Given a empirical distribution $\hat{p}(\ubm{x}) = \frac{1}{R}\sum_{r=1}^{R} \delta_{\ubmr{x}{r}}(\ubm{x})$. We want to find a probabilistic model such that:
\begin{equation}
  \min KL(\hat{p}(\ubm{x})\| p(\ubm{x}))
\end{equation}
where $KL(\cdot\|\cdot)$ denotes the Kullback-Leibler divergence.

@@

\section{Proposal}

\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \tikzstyle{enode} = [thick, draw=blue, ellipse, inner sep = 1pt,  align=center]
    % \tikzstyle{nnode} = [thick, rectangle, rounded corners = 1pt,minimum size = 1.2cm,draw,inner sep = 2pt]
    \tikzstyle{nnode} = [thick, rectangle, rounded corners = 2pt,minimum size = 0.8cm,draw,inner sep = 2pt]
    % \node[enode] (z1) at (-1.2,1.8) {Source 1}; 
    \node[nnode] (g1) at (-0.5,1.8) {Source 1};

    % \node[enode] (z2) at (-1.2,0.5){$\bm{z}\sim p_{2}(\bm{z})$};
    \node[nnode] (g2) at (-0.5,0.5) {Source 2};

    % \node[enode] (zK) at (-1.2,-1.8) {$\bm{z}\sim p_{|\Ss|}(\bm{z})$};
    \node[nnode] (gs) at (-0.5, -1.8) {Source $|\Ss|$};

    \node[enode] (x) at (4.5,1.5){$\ubm{x}\sim p(\ubm{x};\bm{H})$};
    % \node at (3,-1){$p(\bm{z}) = \sum_k\pi_kp_k(\bm{z})$};

    \draw[dotted,line width=2pt] (0,-0.3) -- (0,-1.2);
    \filldraw[->] (1.9, 0.5)circle (2pt) --  (x) ;

    \draw[->] (g1) -- (1.8, 1.8);

    \draw[->] (g2) -- (1.8, 0.5);

    % \draw[->] (1,-0.5) -- (1.8, -0.5);

    \draw[->] (gs) -- (1.8, -1.8);

    % \draw[->] (z1) [in=180,out=0]  to node[right]{$\pi_1$} (g);
    % \draw[->] (z2) [in=180,out=0]  to node[above]{$\pi_2$} (g);
    % \draw[->,dashed] (zK) [in=180,out=0]  to node[right]{$\pi_K$} (g);
    % \draw[->] (g) to (x);
    \begin{scope}[xshift=0.5cm, thick, every node/.style={sloped,allow upside down}]
      \node[nnode] (m) at (3.5,-2) {Memory};
      \node[nnode] (a) at (3.5,-0.5) {$\bm{A}$};

      \draw (2.1,0.9)-- (2.2, 0.);
      \draw (2.2,0.)-- node {\midarrow} (2.2,-2);
      \draw (2.2,-2)-- (m);
      \draw (m)-- (5, -2);
      \draw (5, -2)-- node {\midarrow} (5 ,-0.5);
      \draw (5, -0.5) -- (a);
      \draw (a)-- node {\midarrow} (2.2, -0.5);
      \node at (4.8, -1) {$s_{t}$};
      \node at (2.56, -0.25) {$s_{t+1}$};
    \end{scope}

  \end{tikzpicture}
  \caption{HMM Model defined by $\bm{H} = \{\Ss, \bm{q}, \bm{A}, p({\bm{x}}|{\bm{s}}; \bm{\Phi}) \}$}
  \vspace{0.1cm}
\end{figure}

Since model $\bm{H}$ contains hidden sequential variable $\ubm{s}$, we can not directly solve the maximum likelihood problem in \autoref{eq:ml-of-hmm}. We use expectation maximization (EM) to address the hidden variable problem by
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  The ``expected likelihood'' function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \ \sum_{r=1}^{R}E_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})}\left[\log\,p(\ubmr{x}{r}, \ubmr{s}{r}; \bm{H}) \right]
  \end{equation}
\item M-step: the optimization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}})
  \end{equation}
\end{itemize}

The \autoref{eq:em-m-opt} can be reformulated as:
\begin{equation}\label{eq:m-step-subs}
  \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{{A}}\Qq({A}; \bm{H}^{\mathrm{old}}) + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
\end{equation}
where
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,p(\smtr{s}{1}{r}; \bm{q})\right] \label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})\right] \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,p(\ubmr{x}{r} | \ubmr{s}{r}; \bm{\Phi})\right]\label{eq:generative-model-update}
\end{align}

@@

\begin{algorithm}[H]
  \caption{Meta algorithm for HMM powered by Generative Models (GMHMM)}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:}{
      Building $\bm{H}^{\mathrm{old}}, \bm{H} \in \Hh$ gives: \\
      $\bm{H}^{\mathrm{old}} = \{\Ss, \bm{q}^{\mathrm{old}}, A^{\mathrm{old}}, p(\bm{x}|s; \bm{\Phi}^{\mathrm{old}})\}$, $\bm{H} = \{\Ss, \bm{q}, A, p(\bm{x}|s; \bm{\Phi})\}$
    }, \\
    \STATE Initialize $\bm{H}$
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$
    \FOR { $\bm{H}$ not converge}
    \STATE Sample a batch of data $\left\{ \ubmr{x}{r} \right\}_{r=1}^{R_b}$ from the dataset $\hat{p}(\ubm{x})$
    \STATE Compute $p(\smtr{s}{t}{r} | \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$, $p(\smtr{s}{t}{r}, \smtr{s}{t+1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$ by forward/backward algorithm;
    \STATE $\bm{q} \gets \uargmin{\bm{q}}\, \Qq(\bm{q}; \bm{H}^{\mathrm{old}})$ by \autoref{eq:update-initial-state-prob};
    \STATE $\bm{A} \gets \uargmin{\bm{A}}\Qq(\bm{A}; \bm{H}^{\mathrm{old}})$ by \autoref{eq:update-transition-prob};
    \STATE $\bm{\Phi} \gets \uargmin{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})$ by calling Algorithm \autoref{algo:opt-Phi-GenHMM} or \autoref{algo:opt-Phi-LatHMM};
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$

@@

\subsection{Initial Probability Update}
\autoref{eq:init-distribution-update} can be written as:
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R}\sum_{\ubmr{s}{r}} {p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}; \bm{q}) \nonumber \\
                                     & = \sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}\sum_{\smtr{s}{2}{r}=1}^{|\Ss|}\cdots \sum_{\smtr{s}{T^{r}}{r}}^{{|\Ss|}} {p(\smtr{s}{1}{r}, \smtr{s}{2}{r}, \cdots, \smtr{s}{T^{r}}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}; \bm{q}) \\
                                     & = \sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}{p(\smtr{s}{1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}; \bm{q}) 
\end{align}

@@

\subsection{Transition Probability Update}
\autoref{eq:transition-update} can be written as
\begin{align}
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}})
  &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})\right] \nonumber\\
  &= \sum_{r=1}^{R} \sum_{\ubmr{s}{r}}{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A}) \nonumber \\
  &= \sum_{r=1}^{R} \sum_{t=1}^{T^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}\sum_{\smtr{s}{t+1}{r}=1}^{|\Ss|}{p(\smtr{s}{t}{r}, \smtr{s}{t+1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})
\end{align}

@@

\subsection{Generative Model Update}
\autoref{eq:generative-model-update} can be rewritten as
\begin{align}\label{eq:gm-update}
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
  &= \sum_{r=1}^{R}\sum_{\ubmr{s}{r}}{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\ubmr{x}{r} | \ubmr{s}{r}; \bm{\Phi}) \nonumber \\
  &= \sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}  p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \log\, p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}; \bm{\Phi}).
\end{align}

Then the third subproblem of \autoref{eq:m-step-subs} becomes:
\begin{align}\label{eq:sub-gm}
  &\uargmax{\bm{\Phi}} \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}), \nonumber \\
  \mathrm{s.t.} &\hspace{0.2cm} p(\bm{x} | {s}; \bm{\Phi}) \mathrm{~is~our~ general~model}
\end{align}

It could be seen from \autoref{eq:gm-update} that the key to update generate  model is to evaluate $p(\bm{x}|{s}; \bm{\Phi})$ for all $s \in \Ss$. In Forward/Backward algorithm, evaluation of $p(\bm{x}|{s}; \bm{\Phi})$ is also all what is needed to compute $p(s|\bm{x};\bm{\Phi})$. In the following two subsections, we will provide two neural network based generative models that fulfill this requirement and also have high capability for complex signal modeling.

>>>>

%\def\year{2020}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai20}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{graphicx}  % DO NOT CHANGE THIS
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
%\nocopyright
%PDF Info Is REQUIRED.
% For /Author, add all authors within the parentheses, separated by commas. No accents or commands.
% For /Title, add Title in Mixed Case. No accents or commands. Retain the parentheses.
 \pdfinfo{/Title gm_hmm
} %Leave this	
\newcommand{\citet}[1]{\citeauthor{#1} \shortcite{#1}}
\newcommand{\citep}{\cite}
\newcommand{\citealp}[1]{\citeauthor{#1} \citeyear{#1}}
%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{tikz}
\usetikzlibrary{calc,shapes,positioning}
\usetikzlibrary{arrows}
\newcommand{\midarrow}{\tikz \draw[-triangle 90] (0,0) -- +(.1,0);}
% Be sure to use PDxF Latex
\pdfoutput=1

@@

\usepackage{caption}
\usepackage{bm}
\newcommand{\ubar}[1]{\mkern2mu\underline{\mkern-2mu #1\mkern-2mu}\mkern2mu}
% \allowdisplaybreaks
\usepackage{mystyle}
\newcommand{\ubm}[1]{\ubar{\bm{#1}}}
\newcommand{\ubmr}[2]{\ubar{\bm{#1}}^{(#2)}}

@@

% \hypersetup{  
%   bookmarks=true,
%   backref=true,
%   pagebackref=false,
%   colorlinks=true,
%   linkcolor=blue,
%   citecolor=red,
%   urlcolor=blue,
%   pdftitle={Generative Model HMM},
%   pdfauthor={Dong Liu},
%   pdfsubject={}
% }

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai20.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\setlength\titlebox{2.5in} % If your paper contains an overfull \vbox too high warning at the beginning of the document, use this
% command to correct it. You may not alter the value below 2.5 in

\title{Powering Hidden Markov Model by Generative Models}
\author{
  Dong Liu, Antoine Honoré, Saikat Chatterjee, and Lars~K. Rasmussen\\
  KTH Royal Institute of Technology, Stockholm, Sweden \\
  E-mail: \{doli, honore, sach, lkra\}@kth.se}

@@

\maketitle
\begin{abstract}
to be written...
\end{abstract}

\section{introduction}

\section{Preliminaries}
\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \tikzstyle{enode} = [thick, draw=black, ellipse, inner sep = 1pt,  align=center]
    \tikzstyle{nnode} = [thick, rectangle, rounded corners = 2pt,minimum size = 0.8cm,draw,inner sep = 2pt]
    \node[enode] (g1) at (-0.5,1.8) {$p(\bm{x}| s=1; \bm{\Phi}_{1})$};
    \node[enode] (g2) at (-0.5,0.5) {$p(\bm{x}| s=2; \bm{\Phi}_{2})$};
    \node[enode] (gs) at (-0.5, -1.8) {$p(\bm{x}| s=|\Ss|; \bm{\Phi}_{|\Ss|})$};
    \node[enode] (x) at (4.5,1.5){$\ubm{x}\sim p(\ubm{x};\bm{H})$};

    \draw[dotted,line width=2pt] (0,-0.3) -- (0,-1.2);
    \filldraw[->] (1.9, 0.5)circle (2pt) --  (x) ;
    \draw[->] (g1) -- (1.8, 1.8);
    \draw[->] (g2) -- (1.8, 0.5);
    \draw[->] (gs) -- (1.8, -1.8);

    \begin{scope}[xshift=0.5cm, thick, every node/.style={sloped,allow upside down}]
      \node[nnode] (m) at (3.5,-2) {Memory};
      \node[nnode] (a) at (3.5,-0.5) {$\bm{A}$};

      \draw (2.1,0.9)-- (2.2, 0.);
      \draw (2.2,0.)-- node {\midarrow} (2.2,-2);
      \draw (2.2,-2)-- (m);
      \draw (m)-- (5, -2);
      \draw (5, -2)-- node {\midarrow} (5 ,-0.5);
      \draw (5, -0.5) -- (a);
      \draw (a)-- node {\midarrow} (2.2, -0.5);
      \node at (4.8, -1) {$s_{t}$};
      \node at (2.56, -0.25) {$s_{t+1}$};
    \end{scope}
  \end{tikzpicture}
  \caption{HMM model illustration.}\label{fig:hmm}
  \vspace{0.1cm}
\end{figure}

Our framework is an Hidden Markov Model (HMM). An HMM model $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the signal at time $t$, $[\cdot]^{\intercal}$ means transpose, and $T$ denote the time length. Using HMM for signal representation or generating is illustrated in \ref{fig:hmm}. The assumption is that different instant signal of $\ubar{\bm{x}}$ is generated by different source, while the source selection is done by a discrete hidden markov process. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}$, where:
\begin{itemize}
\item $\Ss$ is the set of states of HMM $\bm{H}$;
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ initial distribution of HMM $\bm{H}$ with $|\Ss|$ is cardinality of $\Ss$, $q_k = p(s=k)$ for random state variable $s$. In the following of this paper, we use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ is the transition matrix for the HMM $\bm{H}$ of size $|\Ss| \times |\Ss|$, $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i)$.
\item For given hidden state $s$, the observable signal density is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this conditional probabilistic model.
\end{itemize}

In the framework of HMM, at each time instance, signal $\bm{x}_t$ is generated by density $p(\bm{x}_t| s_t; \bm{\Phi}_{s_t})$, and $s_t$ is decided by the hidden markov process. This process gives us the probabilistic model $p(\ubm{x};\bm{H})$
\section{Generator-mixed HMM}

\begin{figure}[!th]
  \centering
  \begin{tikzpicture}
    \tikzstyle{enode} = [thick, draw=black, ellipse, inner sep = 2pt,  align=center]
    \tikzstyle{nnode} = [thick, rectangle, rounded corners = 2pt,minimum size = 0.8cm,draw,inner sep = 2pt]
    \node[enode] (z1) at (-1.2,1.8) {$\bm{z}\sim p_{s,1}(\bm{z})$};
    \node[nnode] (g1) at (1,1.8) {$\bm{g}_{s,1}$};
    \node[enode] (z2) at (-1.2,0.5){$\bm{z}\sim p_{s,2}(\bm{z})$};
    \node[nnode] (g2) at (1,0.5) {$\bm{g}_{s,2}$};
    \node[enode] (zK) at (-1.2,-1.8) {$\bm{z}\sim p_{s,K}(\bm{z})$};
    \node[nnode] (gs) at (1, -1.8) {$\bm{g}_{s,K}$};
    \node[enode] (x) at (4.5,0){$\bm{x}\sim p(\bm{x}| s; \bm{\Phi}_{s})$};

    \draw[->] (zK) -- (gs);
    \draw[->] (gs) -- (1.8, -1.8);
  \end{tikzpicture}
  \caption{Source $s$ powered by generator mixed generative model (GenHMM).}
  \vspace{0.1cm}
\end{figure}

In this section, we introduce a neural network based generator-mixed HMM model. We term it as GenHMM. GenHMM has a hidden markov process as typical HMM, while the probabilistic model of GenHMM for each hidden state is induced by mixture of neural network based generators. For $s\in \Ss$, we define the state's probabilistic model in GenHMM as
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\Phi}_{s, \kappa})
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, and $\pi_{s, \kappa} = p(\kappa | \bm{x}, s; \bm{\Phi}_{s, \kappa})$. Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]$,
$\bm{\Phi}_s = \left\{ \pi_{s, \kappa}, \bm{\Phi}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$,

\subsubsection{Generator Mixed HMM (GenHMM)}\label{subsubsec:GenHMM}
In this subsection, we specify the GenHMM model and give the learning algorithm of this model.
Base on the generative model assumption, for each state $s$, the probabilistic generative model is

\subsubsection{Flow structure}

\section{Learning Algorithm of GenHMM}
\subsection{Problem Statement}
Given a empirical distribution $\hat{p}(\ubm{x}) = \frac{1}{R}\sum_{r=1}^{R} \delta_{\ubmr{x}{r}}(\ubm{x})$. We want to find a probabilistic model such that:
\begin{equation}
  \min KL(\hat{p}(\ubm{x})\| p(\ubm{x}))
\end{equation}
where $KL(\cdot\|\cdot)$ denotes the Kullback-Leibler divergence.

@@

\subsection{Proposal}

Since model $\bm{H}$ contains hidden sequential variable $\ubm{s}$, we can not directly solve the maximum likelihood problem in \ref{eq:ml-of-hmm}. We use expectation maximization (EM) to address the hidden variable problem by
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  The ``expected likelihood'' function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \ \sum_{r=1}^{R}E_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})}\left[\log\,p(\ubmr{x}{r}, \ubmr{s}{r}; \bm{H}) \right]
  \end{equation}
\item M-step: the optimization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}})
  \end{equation}
\end{itemize}

The \eqref{eq:em-m-opt} can be reformulated as:
\begin{equation}\label{eq:m-step-subs}
  \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{{A}}\Qq({A}; \bm{H}^{\mathrm{old}}) + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
\end{equation}
where
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,p(\smtr{s}{1}{r}; \bm{q})\right] \label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})\right] \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,p(\ubmr{x}{r} | \ubmr{s}{r}; \bm{\Phi})\right]\label{eq:generative-model-update}
\end{align}

@@

\begin{algorithm}[H]
  \caption{Meta algorithm for HMM powered by Generative Models (GMHMM)}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:}{
      Building $\bm{H}^{\mathrm{old}}, \bm{H} \in \Hh$ gives: \\
      $\bm{H}^{\mathrm{old}} = \{\Ss, \bm{q}^{\mathrm{old}}, A^{\mathrm{old}}, p(\bm{x}|s; \bm{\Phi}^{\mathrm{old}})\}$, $\bm{H} = \{\Ss, \bm{q}, A, p(\bm{x}|s; \bm{\Phi})\}$
    }, \\
    \STATE Initialize $\bm{H}$
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$
    \FOR { $\bm{H}$ not converge}
    \STATE Sample a batch of data $\left\{ \ubmr{x}{r} \right\}_{r=1}^{R_b}$ from the dataset $\hat{p}(\ubm{x})$
    \STATE Compute $p(\smtr{s}{t}{r} | \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$, $p(\smtr{s}{t}{r}, \smtr{s}{t+1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})$ by forward/backward algorithm;
    \STATE $\bm{q} \gets \uargmin{\bm{q}}\, \Qq(\bm{q}; \bm{H}^{\mathrm{old}})$ by \eqref{eq:update-initial-state-prob};
    \STATE $\bm{A} \gets \uargmin{\bm{A}}\Qq(\bm{A}; \bm{H}^{\mathrm{old}})$ by \eqref{eq:update-transition-prob};
    \STATE $\bm{\Phi} \gets \uargmin{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})$ by calling Algorithm~\ref{algo:opt-Phi-GenHMM} or \ref{algo:opt-Phi-LatHMM};
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$

@@

\subsection{Initial Probability Update}
\eqref{eq:init-distribution-update} can be written as:
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R}\sum_{\ubmr{s}{r}} {p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}; \bm{q}) \nonumber \\
                                     & = \sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}\sum_{\smtr{s}{2}{r}=1}^{|\Ss|}\cdots \sum_{\smtr{s}{T^{r}}{r}}^{{|\Ss|}} {p(\smtr{s}{1}{r}, \smtr{s}{2}{r}, \cdots, \smtr{s}{T^{r}}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}; \bm{q}) \\
                                     & = \sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}{p(\smtr{s}{1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}; \bm{q}) 
\end{align}

@@

\subsection{Transition Probability Update}
\eqref{eq:transition-update} can be written as
\begin{align}
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}})
  &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})\right] \nonumber\\
  &= \sum_{r=1}^{R} \sum_{\ubmr{s}{r}}{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A}) \nonumber \\
  &= \sum_{r=1}^{R} \sum_{t=1}^{T^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}\sum_{\smtr{s}{t+1}{r}=1}^{|\Ss|}{p(\smtr{s}{t}{r}, \smtr{s}{t+1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})
\end{align}

@@

\subsection{Generative Model Update}
\eqref{eq:generative-model-update} can be rewritten as
\begin{align}\label{eq:gm-update}
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
  &= \sum_{r=1}^{R}\sum_{\ubmr{s}{r}}{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\ubmr{x}{r} | \ubmr{s}{r}; \bm{\Phi}) \nonumber \\
  &= \sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}  p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \log\, p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}; \bm{\Phi}).
\end{align}

Then the third subproblem of \eqref{eq:m-step-subs} becomes:
\begin{align}\label{eq:sub-gm}
  &\uargmax{\bm{\Phi}} \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}), \nonumber \\
  \mathrm{s.t.} &\hspace{0.2cm} p(\bm{x} | {s}; \bm{\Phi}) \mathrm{~is~our~ general~model}
\end{align}

It could be seen from \eqref{eq:gm-update} that the key to update generate  model is to evaluate $p(\bm{x}|{s}; \bm{\Phi})$ for all $s \in \Ss$. In Forward/Backward algorithm, evaluation of $p(\bm{x}|{s}; \bm{\Phi})$ is also all what is needed to compute $p(s|\bm{x};\bm{\Phi})$. In the following two subsections, we will provide two neural network based generative models that fulfill this requirement and also have high capability for complex signal modeling.
